\begin{center} \bf \sffamily \LARGE Prisoner's dilemma with stochastic payoffs\\[-1cm] \end{center}

### 2023-10-01 paper/main.tex
CH: Polishing the abstract, drafting the intro

\documentclass[11pt]{article}
\usepackage{times}
\usepackage{amsmath,amsthm,amssymb,setspace,enumitem,epsfig,titlesec,verbatim,color,array,eurosym,multirow}
\usepackage[sort&compress]{natbib}
\usepackage[footnotesize,bf]{caption}
\usepackage[margin=2.5cm, includefoot, footskip=30pt]{geometry}
\usepackage{standalone}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{blkarray}
\usepackage[ruled,vlined]{algorithm2e}
\smallskip % Erlaubt kleine Abstaende zwischen Paragraphen, falls es dem Seitenlayout hilft
\renewcommand{\baselinestretch}{1.3}
\newcommand{\R}{\mathbb{R}}

\definecolor{darkblue}{rgb}{0,0,.8}
\definecolor{darkgreen}{rgb}{0,0.5,0.1}
\newcommand{\matlabfunc}[1]{\textcolor{darkblue}{#1}}
\newcommand{\matlabcomment}[1]{\textcolor{darkgreen}{#1}}
\newcommand{\christian}[1]{\textcolor{blue}{\textbf{CH}: #1}}
\newcommand{\alex}[1]{\textcolor{red}{\textbf{AL}: #1}}

@@

\begin{abstract}
\noindent
Models of direct reciprocity explain how cooperation can emerge in repeated social dilemmas. 
According to this literature, individuals should naturally learn to adopt conditionally cooperative strategies if they expect to have further encounters with their interaction partner. 
While models of reciprocity have shaped our understanding of cooperation, they often make idealized assumptions on how individuals remember and process information. 
For example, when strategies change through social learning, it is commonly assumed that individuals take into account the strategy's average payoff.  
However, this would require them to compute (or remember) their payoffs against all other population members.
Instead, individuals may sometimes update their strategies based on their most recent experiences only. 
Herein, we introduce a theoretical framework to study the evolution of reciprocity when individuals only learn based on the outcome of their very last interaction. 
%We compare this model to the classical framework based on expected payoffs.  
We show that cooperation can still evolve, but individuals adopt less generous strategies, and they tend to cooperate less often than in the classical framework based on expected payoffs. 
Interestingly, however, we also show that once individuals remember the payoffs of two or three interactions, evolving cooperation rates approach the results of the classical model. 
These findings contribute to a literature that explores which kind of cognitive capabilities are required for reciprocal cooperation. While our results suggest that more memory facilitates the evolution of reciprocity, it only takes a modest amount of payoff memory for cooperation to emerge. 
\end{abstract}

@@

@@

Evolutionary game theory~\cite{smith1982evolution, hofbauer1998evolutionary,
nowak:Nature:2004, hauert2005game} describes the evolutionary dynamics of
populations consisting of different types of interacting individuals. In the
last two decades the interest of the field has shifted to the analysis of
stochastic finite population dynamics in preference to traditional approaches
such as the replicator dynamics~\cite{hofbauer:JTB:1979, taylor:MATHBIO:1978}.
This is especially true in the topic of cooperation~\cite{hilbe:PNAS:2013,
hilbe:Nature:2018,glynatsi:SCR:2020, ohtsuki:Nature:2006, nowak:Science:2006,
murase:ScientificReports:2020}. In stochastic evolutionary dynamics
disadvantageous mutants have a small yet non-zero probability to reach fixation
and this can lead to fundamental changes in the results. As shown
in~\cite{nowak:Nature:2004}, using such dynamics a single cooperative strategy
can invade a population of defectors without any special modifications in
comparison to traditional approaches that one has to consider spatial
structure~\cite{nowak:Nature:1992} or when payoffs are subject to aggregate
shocks~\cite{fudenberg:JET:1992}.

Stochastic evolutionary dynamics model a finite population. Each
time step of the process consists of three phases; (1) the \textit{mutation
phase} (2) the \textit{game phase} (3) the \textit{update phase}. In the
mutation phase one individual from the population is chosen to switch to a new
mutant strategy with a probability \(\mu\). In the game phase individuals are
randomly matched with other individuals in the population, and they engage in a
repeated game where each subsequent turn occurs with a fixed probability
\(\delta\). The updating phase depends on the process. Two classes of
finite stochastic processes have been used extensively: (i) fitness-based
processes in which an individual chosen proportional to fitness reproduces and
the offspring replaces a randomly chosen individual~\cite{nowak:Nature:2004} or
(ii) \textit{pairwise comparison processes} in which a pair of individuals is
chosen and subsequently one of these individuals may adopt the strategy of the
other~\cite{traulsen2007pairwise}.

Like all theoretical models, finite stochastic processes rest on a
set of assumptions. Namely, in the game phase it is assumed that players use
strategies with finite memory~\cite{Nowak1992tit, Baek2016}. For example, in a
repeated game between two individuals the actions of the players in each turn
are often determined by the action of the co-players in the previous turn~\cite{Nowak1992tit}. This
assumption is common as it allows for an explicit calculation of the players'
payoffs~\cite{sigmund2010calculus}. In addition, it is
assumed that individuals play many times and with all other players before
reproduction takes place. So that in the updating phase the updating payoffs,
or fitness, of individuals is given by the average payoffs of an
infinitely repeated game against every member of the population.

These two assumptions create a curious inconsistency; a player with a finite
memory (in the game stage) can recall infinite information regarding their
interactions and each interaction's outcomes in the update stage. Prior research
has examined the effects of constraining the interactions an individual has.
Either by letting individuals have a stochastic number of
interactions~\cite{roca:PhysicalReview:2006, sanchez:JTB:2005,
Traulsen:JTB:2007} or by only allowing for a single
interaction~\cite{Woelfing:JTB:2009}. However, these studies have not accounted
for the use of strategies with finite memory. Moreover, no previous research has
explored the effect of restricting not only the number of interactions, but also
the amount of information available about each interaction. To this end, we
propose a framework in which individuals, similar to the decisions at each turn,
estimate their fitness based on a minimum of information.

We first consider two extreme scenarios, the classical scenario where an individual
has a perfect memory of all their interactions and the alternative scenario of
limited memory where individuals update their strategies only based on the very
last payoff they obtained. We demonstrate the effects of limited payoff
memory using a pairwise comparison process and the well studied game of the
prisoner's dilemma~\cite{glynatsi:HSSC:2021}. We observe that individuals with
limited payoff memory tend to adopt less generous strategies. We obtain similar
results when we consider that individuals update their strategies based on more
information. More specifically, up to the last two payoffs they obtained when
interacting with up to two different members of the population.

@@

Furthermore, we investigated scenarios where the final two rounds, or the last
two instances of interaction, were taken into account. We observed a
statistically significant rise in the frequency of cooperative behavior. For a
sizable population and a high likelihood of continued interactions, the two
cases yield the same result as the overall payoff is influenced by two possible
outcomes. Notably, the scenario involving two rounds and two interactions
yielded the highest cooperation rate among all the novel methodologies that we
tested.
% A nice sentence missing here.

\section{Acknowledgements}

This work was supported by the European Research Council Starting Grant 850529:
E-DIRECT.

\bibliographystyle{unsrt}
\bibliography{bibliography.bib}

>>>>

\documentclass[11pt]{article}
\usepackage{times}
\usepackage{amsmath,amsthm,amssymb,setspace,enumitem,epsfig,titlesec,verbatim,color,array,eurosym,multirow}
\usepackage[sort&compress,round
]{natbib}
\usepackage[footnotesize,bf]{caption}
\usepackage[margin=2.5cm, includefoot, footskip=30pt]{geometry}
\usepackage{standalone}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{blkarray}
\usepackage[ruled,vlined]{algorithm2e}
\smallskip % Erlaubt kleine Abstaende zwischen Paragraphen, falls es dem Seitenlayout hilft
\renewcommand{\baselinestretch}{1.3}
\newcommand{\R}{\mathbb{R}}

\newcommand{\todo}[1]{\textcolor{blue}{#1}}

@@

\begin{abstract}
\noindent
Direct reciprocity can explain how cooperation emerges in repeated social interactions. 
According to this literature, individuals should naturally learn to adopt conditionally cooperative strategies such as Tit-for-Tat. % if they expect to have further encounters with their interaction partner. 
Corresponding models have greatly facilitated our understanding of cooperation. 
Yet they often make strong assumptions on how individuals remember and process information. 
For example, when strategies are updated through social learning, it is commonly assumed that individuals update their strategies based on their average payoffs.  
This would require them to compute (or remember) their payoffs against all other population members.
Instead, herein, we introduce a theoretical framework to study the evolution of reciprocity when individuals learn based on their most recent experiences.
Even in the most extreme case that they only take into account their very last interaction, we find that cooperation can still evolve. 
However, such individuals adopt less generous strategies, and they tend to cooperate less often than in the classical framework with expected payoffs. 
Interestingly, once individuals remember the payoffs of two or three recent interactions, evolving cooperation rates quickly approach the classical limit. 
These findings contribute to a literature that explores which kind of cognitive capabilities are required for reciprocal cooperation. 
While our results suggest that memory facilitates the evolution of reciprocity, it only takes a rather modest amount of payoff memory for cooperation to emerge. 
\end{abstract}

@@

%%%%%%%%%
%%  INTRO  %%
%%%%%%%%%

@@

%% INTRO: General introduction to evolutionary game theory and cognitive constrains

Evolutionary game theory describes the dynamics of populations when an individual's fitness depends on the traits or strategies of other population members~\cite{hofbauer1998evolutionary, nowak:Nature:2004, hauert2005game}.  
This theory can be used to describe the dynamics of animal conflict~\citep{maynard-smith:Nature:1973}, cancer cells~\citep{Stankova:JAMA:2019}, and of cooperation~\citep{axelrod1981evolution}. 
Respective models translate strategic interactions into games~\cite{smith1982evolution}. 
These games specify how individuals (players) interact, which strategies they can choose, and what fitness consequences (or payoffs) their strategies have. 
In addition, these models also specify the mode by which successful strategies spread over time. 
Models of biological evolution posit that individuals with a high fitness tend to produce more offspring; models of cultural evolution instead assume that such individuals are imitated more often. 
While biological and cultural evolution are sometimes treated as equivalent, there can be important differences~\citep{Wu2015,Smolla:PTRS:2021}. 
For example, models of biological evolution do not require individuals to have any particular cognitive abilities.
Here, it is the evolutionary process itself that biases the population towards strategies with higher fitness. 
In contrast, in models of cultural evolution, individuals need to be aware of the different strategies that are available, and they need to identify those strategies with a higher payoff. 
As a consequence, evolutionary outcomes may often depend on how easily different behaviors can be learned~\citep{Chatterjee:JTB:2012}, and on how easy payoff comparisons are.

%% INTRO: Literature on direct reciprocity and why social learning of strategies is hard

These difficulties -- to learn strategies by social imitation -- are particularly pronounced in models of direct reciprocity. 
This literature follows Trivers' insight that individuals have more of an incentive to cooperate in social dilemmas when they interact repeatedly~\citep{trivers1971evolution}. 
In repeated interactions, individuals can condition their future behavior on their past experiences with their interaction partner. 
They may use strategies such as Tit-for-Tat~\citep{rapoport:book:1965} or Generous Tit-for-Tat~\citep{molander:jcr:1985,Nowak1992tit} to preferentially cooperate with those partners who cooperated in the past. 
Such conditional strategies approximate human behavior fairly well~\citep{fischbacher:EconL:2001,Rand:TCS:2013,DalBo:AER:2019,Rossetti:ETH:2023} and they have also been documented in several other species~\citep{Carter:PRSB:2013,Schweinfurth:AnBehav:2019,Voelkl:PNAS:2015} -- although direct reciprocity is generally more difficult to demonstrate in animals~\citep{CluttonBrock:Nature:2009,Silk:CurrentBiology:2013,taborsky:CurrentBiology:2013}.
However, at the outset, it is not clear how easy it is to {\it learn} such strategies by social imitation. 
As one obstacle, individuals usually only observe other population members' actions (i.e., whether they cooperate or defect). 
These observations alone may not suffice to infer the underlying strategies (i.e., the contingent rules that determine whether to cooperate or defect in a given situation). 
As another obstacle, even if others' strategies are observable, individuals might find it difficult to identify which ones have the highest payoff. 
After all, the payoff of a strategy of direct reciprocity is not determined by the outcome of any single round.
Rather it is determined by how well this strategy fares over an entire sequence of rounds, against many different population members. 
In practice, such information might be both difficult to obtain and difficult to process.

%% INTRO: Gap in the literature and what  question we wish to address.

Most models of direct reciprocity abstract from these difficulties~\citep{brauchli:JTB:1999,brandt:JTB:2006,ohtsuki:JTB:2007b,szolnoki:pre:2009b,imhof2010stochastic,van-segbroeck:prl:2012,grujic:jtb:2012,Martinez2012,stewart:pnas:2013,pinheiro:PLoSCB:2014,stewart:games:2015,Baek2016,McAvoy:ProcA:2019,glynatsi:SCR:2020,Schmid:PlosCB:2022,Murase:SciRep:2022}. 
Instead they assume individuals can easily copy the strategies of others. 
Similarly, they assume that updating decisions are based on the strategies' average (or expected) payoffs, taking into account all rounds and interactions. 
These assumptions create a curious inconsistency in how models represent an individual's cognitive abilities. 
On the one hand, when playing the game, individuals are often assumed to have restricted memory. 
Respective studies typically assume that individuals make their decisions each round based on the outcome of the last round only~\citep[with only a few exceptions, see Refs.][]{Hauert1997,van-veelen:PNAS:2012,Stewart2016,Li:NatCS:2022,Murase:PLoSCompBio:2023a}. 
Yet when learning new strategies, individuals are assumed to remember (or compute) each others' precise average payoff across many rounds and many interaction partners. 
Herein, we wish to explore whether this latter assumption is actually necessary for the evolution of reciprocity through social imitation. 
We ask whether individuals can learn to adopt reciprocal strategies even when learning only based on payoff information from a limited number of rounds.

%% INTRO: Basic description of our model and our findings

To explore that question, we first consider two extreme scenarios. 
The first scenario corresponds to the usual modeling approach. 
Here, individuals update their strategies for a repeated prisoner's dilemma based on pairwise comparisons~\cite{traulsen2007pairwise}, based on their strategies' expected payoff. 
We contrast this model with an alternative scenario where individuals update their strategies based on the very
last (one-shot) payoff they obtained. 
We observe that individuals with limited payoff memory tend to adopt less generous strategies, yet moderate levels of cooperation can still evolve. 
Moreover, as we increase the individuals' payoff memory, to include the last two or three one-shot payoffs, cooperation rates quickly approach the rates observed in the classical baseline case. 
Overall, these findings suggest that while memory is important, already minimal payoff information may suffice for the evolution of direct reciprocity based on social learning. 
They also suggest that the classical model of reciprocity (based on expected payoffs) can be interpreted as a useful approximation to more realistic models that provide a more realistic depiction of cognitive constraints.

@@

Furthermore, we investigated scenarios where the final two rounds, or the last
two instances of interaction, were taken into account. We observed a
statistically significant rise in the frequency of cooperative behavior. For a
sizable population and a high likelihood of continued interactions, the two
cases yield the same result as the overall payoff is influenced by two possible
outcomes. Notably, the scenario involving two rounds and two interactions
yielded the highest cooperation rate among all the novel methodologies that we
tested.\\
% A nice sentence missing here.

%%%%%%%%%%%
%% END NOTES %%
%%%%%%%%%%%

\begin{comment}

\noindent
{\bf Ethics.}
This work is purely theoretical and did not require ethical approval from a human subject or animal welfare committee.\\

\noindent
{\bf Data accessibility.}
All data and code used in this manuscript are openly available at https://github.com/Nikoleta-v3/pd-with-stochastic-payoffs \todo{(Correct?)}.\\

\noindent
{\bf Declaration of AI use.}
We have not used AI-assisted technologies in creating this article.\\

\noindent
{\bf Authors' contributions.}
N.G.: conceptualization, formal analysis, investigation, methodology, visualization, writing - original draft, writing -- review and editing; 
A.M.: conceptualization, formal analysis, methodology, writing -- review and editing; 
C.H.: conceptualization, formal analysis, methodology, writing -- review and editing.\\

\noindent
{\bf Conflict of interest declaration.}
We declare we have no competing interests.\\

\noindent
{\bf Funding.}
N.G. and C.H. acknowledge generous support by the European Research Council Starting Grant 850529:
E-DIRECT, and by the Max Planck Society.\\
\end{comment}

{
{\setlength{\bibsep}{0\baselineskip}\small
\bibliographystyle{naturemag}
\bibliography{bibliography.bib}
}
